---
title: "Temnos2021Analysis"
author: "Phoebe Koenig"
date: "2023-03-09"
output: html_document
---

# Loading

### Packages

```{r}
library(lme4)
library(nlme)
library(car)
library(ggplot2)
library(plyr)
library(dplyr)
library(MASS)
library(factoextra)
library(performance)
library(patchwork)
library(emmeans)
library(MuMIn)
library(rsq)
```

### Data
Import dataset and remove ants without measured venom
```{r}
datastore <- read.csv("~/Temnos2021/Koenig_TemnosVenom_2021.csv")

data<-na.omit(datastore)
```

# Variables

### List:

Dataset has dissected workers and queens. It also has some pupae. Here are the variables in the dataset:

**Date.Collected.D.M** : (DD-Month-YYYY) Date on which colony was collected. Year should always be 2021.

**Date.Dissected.D.M** : (DD-Month-YYYY) Date on which ant was dissected to measure venom sac. Year should be 2021 or 2022.

**Date.Collected.Year** : (YYYY) Year in which colony was collected. Should be 2021 for every colony.

**Colony** : The colony number identifier. Technically, this is a nest, not a colony.

**Adult.Count** : The number of adults in the nest on the day it was censused

**Queen.Count** : The number of dealate queens (wingless) in the nest on the day it was censused

**Larvae.Count** : The number of larvae in the nest on the day it was censused. 

**Pupae.Count** : The number of pupae in the nest on the day it was censused Males.Count : The number of males in the nest on the day it was censused

**Winged.Queen.Count** : The number of alate queens (winged) in the nest on the day it was censused

**Egg.Count** : The number of eggs in the nest on the day it was censused. This number could be an underestimate as eggs are small and hard to count

**Worker.Queen** : Whether the dissected ant was a worker or a queen

**Behavior.Code** : If Worker.Queen=="Worker", this field should be either Nurse or Forager. If Worker.Queen=="Queen", this field should be either Mated or Unmated. Mated means the queen was counted in the Queen.Count and was dealate. Unmated means the queen was either counted in Winged.Queen.Count or hatched while the colony was in the lab, and was alate. For both workers and queens, we also have the option pupa, which is where I dissected a pupa. There are not very many of them and I exclude these data for analyses.

**Callow** : (Binary) was the individual dissected callow? 1 if yes, 0 if not

**Webers.Length.mm** : Weber's length in mm

**Venom.Sac.Length.mm** : Length of venom sac in mm

**Venom.Sac.Width.mm** : Width of venom sac in mm

**Ant** : Ant number (identifier)

**Sac.Collected** : (binary) whether or not the sac was successfully harvested to store in freezer for chemical analysis? 1 if yes, 0 if no

### Add and mutate:

In this chunk, I import the summary datasheet (nest level data) and add a few new variables to the dissection dataset that can be calculated from the initial variables. Almost all info included in the summary dataset is included in the original dataset, but I import the summary dataset in order to scale the nest census variables, because I want them scaled by the average colony size. I also make sure the date is in the correct format. The most important variable created here is **Venom.Volume**, which is calculated with the formula Venom.Volume $=\frac{\pi}{6} \times(L \times W^2)$ for each venom sac. 

I use Date to create a new variable, **Collection.Day** where day 1 is the first day I collected ants in the field. 

I subset the dataset to create separate datasets of workers and queens.

```{r}
library(lubridate)
data$Date.Collected.D.M <- as.Date((data$Date.Collected.D.M), format = "%d-%b-%y")
data$Date.Dissected.D.M <- as.Date((data$Date.Dissected.D.M), format = "%d-%b-%y")
data$Callow<-as.logical(data$Callow)
summarydata<- read.csv("~/Temnos2021/Koenig_Temnos_2021_nest_census_data.csv")
summarydata$Date.Collected <- as.Date(as.character(summarydata$Date.Collected), format = "%d- %b")
year(summarydata$Date.Collected) <- 2021
#R auto assigned current year, but these nests were all collected in 2021, so I reassign year

#In R, first day is automatically Jan 1, 1970. therefore we change to day of collection to make results more interpretable. To do this I first convert Collection date to a numeric, where day 1 is the first day I collected ant nests (since first day in R is 1970 I convert to a numeric and then subtract every day before that, which is 18750 days)
data$Collection.Day<-as.numeric(data$Date.Collected.D.M)-18750


#Scale variables to make results more interpretable. Since there are multiple rows of data in our main dataset for the same nest, we don't want nests that had more workers to contribute more to the mean for adult count or larva count. We want only one contribution to the mean per nest. To do this I pulled in the summarydata dataset which is all colonies collected over the season (some of which never led to successful dissections of workers). I then subset the dataset to make sure I'm only calculating means from nests that did see successful dissections
# subset row by colony numbers that are included in data df
summarydatasubset<-subset(summarydata, summarydata$Colony %in% data$Colony)

#calculate means and SDs for Adult, pupa, and larva count. Define scaled variables in data df 
meanadults<-mean(summarydatasubset$Adult.Count)
SDadults<-sd(summarydatasubset$Adult.Count)
data$Adult.Count.Scaled<-(data$Adult.Count-meanadults)/SDadults
meanpupae<-mean(summarydatasubset$Pupae.Count)
SDpupae<-sd(summarydatasubset$Pupae.Count)
data$Pupae.Count.Scaled<-(data$Pupae.Count-meanpupae)/SDpupae
meanlarvae<-mean(summarydatasubset$Larvae.Count)
SDlarvae<-sd(summarydatasubset$Larvae.Count)
data$Larvae.Count.Scaled<-(data$Larvae.Count-meanlarvae)/SDlarvae
data$Location.Collected<-as.factor(data$Location.Collected)


#Create a variable for venom volume, which is an ellipsoid using venom sac length and width. Since both length and width are measured in mm, this volume is in mm^3, which is 1:1 with microliters. For the purposes of graphing and modeling this problem, we will multiply by 1000 to convert to nanoliters
data$Venom.Volume<-(pi/6)*data$Venom.Sac.Length.mm*data$Venom.Sac.Width.mm^2*1000



#Standardize venom by body size
data$standardizedvenom<-((data$Venom.Volume)/(data$Webers.Length.mm))



#Create a dataset for just workers. At this point these datasets still contain pupae, which must be removed for analyses
workerdata<-subset(data,Worker.Queen=="Worker")
queendata<-subset(data,Worker.Queen=="Queen")
```

### Dataset size

```{r}
nrow(subset(workerdata,Behavior.Code=="Pupa"))
#Dissected 15 worker pupae
nrow(subset(queendata,Behavior.Code=="Pupa"))
#Dissected 9 queen pupae
length(unique(workerdata[["Colony"]]))
length(unique(queendata[["Colony"]]))
#Remove pupae from datasets!!!
workerdata<-subset(workerdata,Behavior.Code!="Pupa")
queendata<-subset(queendata,Behavior.Code!="Pupa")
nrow(workerdata)
#Successfully dissected 2054 adult worker
nrow(subset(workerdata,Behavior.Code=="Nurse"))
length(unique(subset(workerdata,Behavior.Code=="Nurse")$Colony))
nrow(subset(workerdata,Behavior.Code=="Forager"))
length(unique(subset(workerdata,Behavior.Code=="Forager")$Colony))
#1394 of these were nurses from 153 colonies, 661 of them foragers from 142 colonies
nrow(queendata)
#Successfully dissected 199 queens
nrow(subset(queendata,Behavior.Code=="Mated"))
nrow(subset(queendata,Behavior.Code=="Unmated"))
#155 of them were dealate queens, 44 were alate queens. They are coded in dataset as mated or unmated but I did not dissect spermathecas, just looked at whether or not they had wings

nrow(subset(workerdata,Callow==T))
#16
length(unique(subset(workerdata,Callow==T)$Colony))
# from 8 colonies
nrow(subset(workerdata,Callow==F))
length(unique(subset(workerdata,Callow==F)$Colony))
#2039 from 155 colonies

mean(workerdata$Venom.Volume)
sd(workerdata$Venom.Volume)

#What is the average number of nests per collection day?
uniquedays<-unique(workerdata$Collection.Day)
length(uniquedays)
uniquenests<-unique(workerdata$Colony)
length(uniquenests)
length(uniquenests)/length(uniquedays)
#3.9 nests on average per day, 40 collection days


queened<-subset(workerdata,Queen.Count>0)
notqueened<-subset(workerdata,Queen.Count==0)
length(unique(queened$Colony))
length(unique(notqueened$Colony))
```

# Visualizing worker data

### Behavior Code

At this point, we need to remove pupae from dataset.

```{r}
workerdata<-subset(workerdata,Behavior.Code!="Pupa")
queendata<-subset(queendata,Behavior.Code!="Pupa")

```

Do nurses and foragers have the same amount of venom? Do workers in queenright colonies have the same amount of venom as workers in queenless colonies? Do callow workers have the same amount of venom as melanized workers? Graph it.

```{r}

venombybehavior<-ggplot(workerdata,aes(Behavior.Code,Venom.Volume,color=Behavior.Code))+
geom_violin()+
  geom_boxplot(width=0.1)+
  xlab("Behavioral Group")+ylab("Venom Volume (nL)")+theme_bw(base_size = 22)+
  theme(legend.position="none")+
scale_colour_discrete("")

venombyqueenright<-ggplot(workerdata,aes(as.factor(Queen.Count>0),Venom.Volume, color=as.factor(Queen.Count>0)))+
geom_violin()+
  geom_boxplot(width=0.1)+
  xlab("Number of Dealate Queens")+ylab("Venom Volume (nL)")+theme_bw(base_size = 22)+
  theme(legend.position="none")+
  scale_x_discrete(labels = c('0','At least 1'))

callowplot = ggplot(data=workerdata,aes(x=Callow, y=Venom.Volume, color=Callow)) + geom_violin() +geom_boxplot(width=0.1)+
xlab("Callow")+ylab("Venom Volume (nL)")+theme_bw(base_size = 22)+theme(legend.position = "none")+ scale_x_discrete(labels=c("No", "Yes"))

remove_y <- theme(
  axis.text.y = element_blank(),
  axis.ticks.y = element_blank(),
  axis.title.y = element_blank()
)

p=list(venombybehavior,callowplot+remove_y,venombyqueenright+remove_y)

wrap_plots(p, nrow = 1) 

ggsave('workers2021.png',plot=wrap_plots(p, nrow = 1),dpi=900,width =39, height = 15, units = "cm",device='png')

```



Worker venom volume represented graphically

```{r}
#now the initial model, which looks for an interaction effect between behavior and pupae count, and between behavior and adult count.

#exclude dissected pupae from dataset (we already did above, but no problem to have code 2x in case do not run above chunk)
workerdata<-subset(workerdata, Behavior.Code!="Pupa")

hist(workerdata$Venom.Volume)
#A bit right skewed

callowplot = ggplot(data=workerdata,aes(x=Callow, y=standardizedvenom, color=Callow)) + geom_violin() +geom_boxplot(width=0.1)+
xlab("Callow")+ylab("Standardized Venom Volume")+theme_bw(base_size = 18)+theme(legend.position = "none")+ scale_x_discrete(labels=c("No", "Yes"))
callowplot

venombysize = ggplot(data=workerdata,aes(x=Webers.Length.mm, y=Venom.Volume)) + geom_point() +
  xlab("Weber's Length (mm)")+ylab("Venom Volume")+theme_bw(base_size = 22)+geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE)
venombysize


  
venombydate = ggplot(data=workerdata, aes(Date.Collected.D.M,Venom.Volume)) + 
geom_point() +
  scale_x_date(date_breaks = "months" , date_labels = "%b")+
geom_smooth(method=lm, se=TRUE ) +
xlab("Date of Nest Collection ")+
ylab("Venom Volume (nL)")+
theme_bw(base_size = 22)
venombydate
ggsave('venombydate2021revised.png',plot=venombydate,dpi=900,width =25, height = 15, units = "cm",device='png')

venombypupae = ggplot(data=workerdata, aes(Pupae.Count,standardizedvenom)) + 
geom_point() +
geom_smooth(method=lm, color="red", fill="gray", se=TRUE) +
xlab("Number of Pupae in Nest")+
ylab("Standardized Venom Volume")+
theme_bw(base_size = 22)
venombypupae

venombyadultinteraction=ggplot(data=workerdata, aes(Adult.Count,Venom.Volume, color=Behavior.Code)) + 
geom_point() +
geom_smooth(method=lm, fill="gray", se=TRUE) +
xlab("Number of Workers in Nest")+
ylab("Venom Volume (nL)")+
theme_bw(base_size = 22)+
scale_colour_discrete("Behavior")
venombyadultinteraction
ggsave('adultinteraction2021revised.png',plot=venombyadultinteraction,dpi=900,width =30, height = 15, units = "cm",device='png')


#Define ratio of pupae to adults
workerdata$Pupae.Adult.Ratio=workerdata$Pupae.Count/workerdata$Adult.Count

venombyratio=ggplot(data=workerdata, aes(Pupae.Adult.Ratio,Venom.Volume, color=Behavior.Code)) + 
geom_point() +
geom_smooth(method=lm, fill="gray", se=TRUE) +
xlab("Ratio of Pupae to Adults")+
ylab("Venom Volume (nL)")+
theme_bw(base_size = 22)+
scale_colour_discrete("Behavior")
venombyratio
ggsave('pupaeinteraction2021revised.png',plot=venombyratio,dpi=900,width =30, height = 15, units = "cm",device='png')
```

# Modeling Worker data

### by worker

Next, lets run a generalized linear mixed-effects model

```{r}
workerdata$Pupae.Adult.Ratio=workerdata$Pupae.Count/workerdata$Adult.Count
workerdata$Larvae.Adult.Ratio=workerdata$Larvae.Count/workerdata$Adult.Count
workerdata$Young.Adult.Ratio=(workerdata$Larvae.Count+workerdata$Pupae.Count)/workerdata$Adult.Count

model1<-lme(Venom.Volume~Callow+Collection.Day+Behavior.Code*Adult.Count.Scaled+Behavior.Code*Pupae.Adult.Ratio+as.factor(Queen.Count>0)+Larvae.Adult.Ratio+Winged.Queen.Count+I(Webers.Length.mm^3), data=workerdata, random= ~1|Colony)
summary(model1)
emtrends(model1,pairwise~Behavior.Code, var = "Pupae.Adult.Ratio",infer=TRUE)
emtrends(model1,pairwise~Behavior.Code, var = "Adult.Count.Scaled",infer=TRUE)
#Remove Larvae.Adult.Ratio from model

model2<-lme(Venom.Volume~Callow+Collection.Day+Behavior.Code*Adult.Count.Scaled+Behavior.Code*Pupae.Adult.Ratio+as.factor(Queen.Count>0)+Winged.Queen.Count, data=workerdata, random= ~1|Colony)
summary(model2)
emtrends(model2,pairwise~Behavior.Code, var = "Pupae.Adult.Ratio",infer=TRUE)
emtrends(model2,pairwise~Behavior.Code, var = "Adult.Count.Scaled",infer=TRUE)
#Remove Winged.Queen.Count from model

model3<-lme(Venom.Volume~Callow+Behavior.Code*Adult.Count.Scaled+Collection.Day+Behavior.Code*Pupae.Adult.Ratio+as.factor(Queen.Count>0)+I(Webers.Length.mm^3), data=workerdata, random= ~1|Colony)
summary(model3)
emtrends(model3,pairwise~Behavior.Code, var = "Pupae.Adult.Ratio",infer=TRUE)
emtrends(model3,pairwise~Behavior.Code, var = "Adult.Count.Scaled",infer=TRUE)
#Callow workers have significantly less venom
#Workers have less venom as season progresses from may to october
#significant interaction effect between behavior and adult count
#significant interaction effect between behavior and pupae count

plot(model3)
qqnorm(resid(model3))
hist(resid(model3))

#Overall, slopes are significantly different from each other. MEANING that the slope for nurses to predict venom based on adult count is significantly different than that of foragers, and same for pupae. 
#For Pupae adult ratio, slope for nurses is the only one that is significantly different than 0. So as the number of pupae in a colony increase, the nurses in that colony have more venom.
#This is another way to do an ad hoc test below. Basically, this code says at -1,0,and 1, slopes for nurses and foragers are significantly different from each other. Here it is only written for worker count
pairs(emmeans(model3, ~ Behavior.Code * Adult.Count.Scaled,at=list(Adult.Count.Scaled=c(-1,0,1))))
pairs(emmeans(model3, ~ Behavior.Code | Adult.Count.Scaled,at=list(Adult.Count.Scaled=c(-1,0,1))))


rsq.lmm(model3)

```

In this chunk, I am using the performance package function check_collinearity to test if date and pupae are collinear (and a problem), since pupae are only existant at certain points during the season

```{r}
check_collinearity(model3)
#No problems here! All well under 5, and even under 3.
```

###Check Outliers
We found a few nests with very high pupae to worker ratios. If we remove these outliers, do our results stay the same?
```{r}
#What the figure looks like with outliers removed (Pupae.Adult.Ratio>=5 removed)
venombyratio2=ggplot(data=subset(workerdata,Pupae.Adult.Ratio<5), aes(Pupae.Adult.Ratio,standardizedvenom, color=Behavior.Code)) + 
geom_point() +
geom_smooth(method=lm, fill="gray", se=TRUE) +
xlab("Ratio of Pupae to Adults")+
ylab("Standardized Venom Volume")+
theme_bw(base_size = 22)+
scale_colour_discrete("Behavior")
venombyratio2

#Model results with removed
model3_OutliersRemoved<-lme(Venom.Volume~Callow+Behavior.Code*Adult.Count.Scaled+Collection.Day+Behavior.Code*Pupae.Adult.Ratio+as.factor(Queen.Count>0)+I(Webers.Length.mm^3), data=subset(workerdata,Pupae.Adult.Ratio<5), random= ~1|Colony)
summary(model3_OutliersRemoved)
emtrends(model3_OutliersRemoved,pairwise~Behavior.Code, var = "Pupae.Adult.Ratio",infer=TRUE)
emtrends(model3_OutliersRemoved,pairwise~Behavior.Code, var = "Adult.Count.Scaled",infer=TRUE)

#Nurses from colonies with higher pupae to adult ratios still have more venom even when we remove these outliers

```



###Controlling for Size of ants
Which is the best way to control for Weber's length? Do we standardize response variable or include it in the model? Do we cube it or leave it linear? Some of this code from Lynn Johnson at cornell statistical consulting unit
```{r}
plot(workerdata$standardizedvenom, workerdata$Venom.Volume/workerdata$Webers.Length.mm)
workerdata$standardizedvenom2 <- workerdata$Venom.Volume/(workerdata$Webers.Length.mm^3)
workerdata$Pupae.Adult.Ratio <- workerdata$Pupae.Count/workerdata$Adult.Count


# venom/weber
model3 <- lme(standardizedvenom ~ Callow+Behavior.Code*Adult.Count.Scaled+Collection.Day+
              Behavior.Code*Pupae.Adult.Ratio+as.factor(Queen.Count>0), random= ~1|Colony,
              data=workerdata)
summary(model3)
emtrends(model3, pairwise~Behavior.Code, var = "Pupae.Adult.Ratio",infer=TRUE)
emtrends(model3, pairwise~Behavior.Code, var = "Adult.Count.Scaled",infer=TRUE)

# venom ~ weber
model3_fe <- lme(Venom.Volume ~ Callow+Behavior.Code*Adult.Count.Scaled+Collection.Day+
                       Behavior.Code*Pupae.Adult.Ratio+as.factor(Queen.Count>0) + Webers.Length.mm, 
                     random= ~1|Colony, data=workerdata)
summary(model3_fe)
emtrends(model3_fe, pairwise~Behavior.Code, var = "Pupae.Adult.Ratio",infer=TRUE)
emtrends(model3_fe, pairwise~Behavior.Code, var = "Adult.Count.Scaled",infer=TRUE)

# venom/weber^3
model3_cubed <- lme(standardizedvenom2 ~ Callow+Behavior.Code*Adult.Count.Scaled+Collection.Day+
              Behavior.Code*Pupae.Adult.Ratio+as.factor(Queen.Count>0), random= ~1|Colony, data=workerdata)
summary(model3_cubed)
emtrends(model3_cubed, pairwise~Behavior.Code, var = "Pupae.Adult.Ratio",infer=TRUE)
emtrends(model3_cubed, pairwise~Behavior.Code, var = "Adult.Count.Scaled",infer=TRUE)

# venom ~ weber^3
model3_cubed_fe <- lme(Venom.Volume ~ Callow+Behavior.Code*Adult.Count.Scaled+Collection.Day+
              Behavior.Code*Pupae.Adult.Ratio+as.factor(Queen.Count>0) + I(Webers.Length.mm^3), 
              random= ~1|Colony, data=workerdata)
summary(model3_cubed_fe)
emtrends(model3_cubed_fe, pairwise~Behavior.Code, var = "Pupae.Adult.Ratio",infer=TRUE)
emtrends(model3_cubed_fe, pairwise~Behavior.Code, var = "Adult.Count.Scaled",infer=TRUE)



plot(workerdata$standardizedvenom, fitted(model3), xlim = c(0,0.0035), ylim = c(0,0.0035))
cor(workerdata$standardizedvenom, fitted(model3))

plot(workerdata$Venom.Volume, fitted(model3_fe), xlim = c(0,0.0035), ylim = c(0,0.0035))
cor(workerdata$Venom.Volume, fitted(model3_fe))

plot(workerdata$standardizedvenom2, fitted(model3_cubed), xlim = c(0,0.01), ylim = c(0,0.01))
cor(workerdata$standardizedvenom2, fitted(model3_cubed))

plot(workerdata$Venom.Volume, fitted(model3_cubed_fe), xlim = c(0,0.004), ylim = c(0,0.004))
cor(workerdata$Venom.Volume, fitted(model3_cubed_fe))



```



### Breakpoint

Investigating the idea of a breakpoint analysis. Ultimately decided to use date as a continuous variable because it's hard to determine how to bin date.

```{r}
#Pupae first in colonies on june 14. Almost none after august 20

for (row in 1:nrow(workerdata)) {
    
    if(workerdata$Date.Collected.D.M[row]<"2021-06-14") {
        workerdata$Raid.Potential[row]="too early"
        }else if(workerdata$Date.Collected.D.M[row]>"2021-08-20"){workerdata$Raid.Potential[row]="too late"
  }else{workerdata$Raid.Potential[row]="raids possible"}
}
workerdata$Raid.Potential <- factor(workerdata$Raid.Potential, levels = c("too early", "raids possible", "too late"))
breakpointmodel<-lme(standardizedvenom~Callow+Raid.Potential+Behavior.Code*Adult.Count.Scaled+Behavior.Code*Pupae.Adult.Ratio+as.factor(Queen.Count>0), data=workerdata, random= ~1|Colony)
summary(breakpointmodel)
emtrends(breakpointmodel,pairwise~Behavior.Code, var = "Adult.Count.Scaled",infer=TRUE)
emtrends(breakpointmodel,pairwise~Behavior.Code, var = "Pupae.Adult.Ratio",infer=TRUE)
emmeans(breakpointmodel, specs = pairwise ~ Raid.Potential)


venombyraidpotential<-ggplot(workerdata,aes(Raid.Potential,standardizedvenom),color=Behavior.Code)+
  geom_boxplot()+
  xlab("Raid Potential")+ylab("Standardized Venom Volume")+theme_bw(base_size = 14)+
  theme(legend.position="none")+
scale_colour_discrete("")
venombyraidpotential


workerdata$Raid.Potential <- factor(workerdata$Raid.Potential, levels = c("early", "maximum potential", "too late"))
for (row in 1:nrow(workerdata)) {
    
    if(workerdata$Date.Collected.D.M[row] <"2021-06-15") {
        workerdata$Raid.Potential[row]="early"
    }
  else if (workerdata$Date.Collected.D.M[row]>"2021-07-20"){workerdata$Raid.Potential[row]="too late"}
  else {workerdata$Raid.Potential[row]="maximum potential"}
}


maxraidpotentialmodel<-lme(standardizedvenom~Callow+Raid.Potential*Behavior.Code+Behavior.Code*Adult.Count.Scaled+Behavior.Code*Pupae.Adult.Ratio+as.factor(Queen.Count>0), data=workerdata, random= ~1|Colony)
summary(maxraidpotentialmodel)
emtrends(maxraidpotentialmodel,pairwise~Behavior.Code, var = "Adult.Count.Scaled",infer=TRUE)
emtrends(maxraidpotentialmodel,pairwise~Behavior.Code, var = "Pupae.Adult.Ratio",infer=TRUE)
emmeans(maxraidpotentialmodel, specs = pairwise ~ Raid.Potential|Behavior.Code)


```


# Nest data

```{r}
#Since dataset doesn't have year in same column, we have to jump through some hoops to get this all set up



par(mfrow=c(2,1))
plot(summarydata$Date.Collected,summarydata$Pupae.Count, xlab="Date of Colony Collection", ylab="Pupae Count")

g = ggplot(data=summarydata, aes(x=Date.Collected)) + geom_point(aes(y=Adult.Count),color="red") + geom_point(aes(y=Pupae.Count),color="blue")+ geom_point(aes(y=Larvae.Count),color="orange")+
xlab("Date Collected")+ylab("Nest Census Count")+
theme_bw(base_size = 22)

summarydata$bin <- cut(summarydata$Date.Collected, breaks=c("months"))
ggplot(data=summarydata) + geom_boxplot(aes(bin, Adult.Count),color="red")+ geom_boxplot(aes(bin, Pupae.Count),color="blue")


dat <- reshape2::melt(summarydata, id.vars="Date.Collected", 
                      measure.vars=c("Adult.Count", "Larvae.Count","Pupae.Count"))
dat$bin <- cut(dat$Date.Collected, breaks=c("months"))

g<-ggplot(data=dat) + geom_boxplot(aes(bin, value, fill=variable))+
xlab("Date Collected")+
ylab("Nest Census")+
theme_bw(base_size = 20)+
scale_x_discrete(labels=c("May","June","July","Aug","Sep","Oct"))+
theme(legend.position = c(0.095, 0.85),legend.background = element_rect(fill = "white"))+
scale_fill_brewer(palette="BuPu",name = "Ant Type", labels = c("Workers", "Larva", "Pupa"))+
  theme(legend.background = element_rect(colour = 'black', fill = 'white', linetype='solid'))



queen=ggplot(data=summarydata, aes(x=Date.Collected, y=Queen.Count)) + geom_point() +
xlab("Date Collected")+ylab("Number of Dealate Queens")+
theme_bw(base_size = 22)+
scale_x_date(date_breaks = "months" , date_labels = "%b")+
geom_smooth(method='lm', formula= y~x)


summarydata$Collection.Day<-as.numeric(summarydata$Date.Collected)-18750

censusadult<-lme(Adult.Count~bin, data=summarydata, random= ~1|Colony)
summary(censusadult)
#Adult count significantly increases with date, aug/sep/oct significantly higher than may/june/july
censusadult<-lme(Adult.Count~Collection.Day, data=summarydata, random= ~1|Colony)
summary(censusadult)




```

